# Reproducible Research: Peer Assessment 1

This report was created to fulfill Course Project 1 in the 'Reproducible Research' course in the Data Science Signature Track for Coursera.

It was completed on 5/12/2014.

tim@timmarco.com


## Dependencies
This script requires the lattice package for outputting graphics.

```{r}

library(lattice)


```


## Loading and preprocessing the data


```{r}

raw.data <- read.csv("activity.csv")

```



## What is mean total number of steps taken per day?

First, we calculate the **total** number of steps on a given date. A histogram of these results can be seen below:

```{r}
total.steps <- aggregate(raw.data$step ~ raw.data$date, data=raw.data, sum, na.rm = FALSE)

hist(total.steps[,2], xlab="Number of Total Steps", ylab="Frequency (Number of Days)", main="Histogram of Total Observed Steps by Date")
```


Next, we calculate the **mean** and **median** number of steps on a given date. Values by date are shown below:


### Mean

```{r}
mean.steps.per.day <- mean(total.steps[,2])
mean.steps.per.day
```



### Median

```{r}
median.steps.per.day <- median(total.steps[,2])
median.steps.per.day
```

The mean value for number of steps per day is **10,766**. The median value is very close: **10,765**. This data set appears to be quite normally distributed.

## What is the average daily activity pattern?

In this step, we will calculate the **average number of steps** taken, aggregated by the five-minute intervals in the original data set. The line chart below plots these values.

*Note: 'average' here is interpreted to mean the 'mean' value, not the median.*


```{r}

average.steps.by.interval <- aggregate(raw.data$step ~ raw.data$interval, data=raw.data, mean, na.rm = FALSE)
names(average.steps.by.interval) <- c("interval","steps")
plot(average.steps.by.interval,type="l", xlab="Five-Minute Interval", ylab="Average Number of Steps", main="Average number of Steps by Five Minute Interval")


```


Next, we get the **interval with the highest average steps**. This is stored as *max.interval*.

```{r}

max.interval <- average.steps.by.interval[average.steps.by.interval$steps ==  max(average.steps.by.interval$steps),1]
max.interval
```

The interval with the highest average number of steps is **835**.




## Imputing missing values

First, we count the **number of NA values in the data set**. This value is stored as *count.raw*.

```{r}

count.raw <- sum(is.na(raw.data))
count.raw

```

NA values from the original dataset are replaced by the average for the appropriate interval. These values are taken from the *average.steps.by.interval* data.frame. These are stored 



```{r}

replace.nas <- function(row) {
  if(is.na(row[1])) {
    current.interval <- as.numeric(row[3])
    new.value <- average.steps.by.interval[average.steps.by.interval$interval == current.interval,]
    new.value <- new.value[2]
  } else {
    new.value <- row[1]
  }
  
  as.numeric(new.value)
  new.value[1]
}


cleaned.data <- raw.data

cleaned.data$steps <- apply(cleaned.data,1,replace.nas)


```




### Histogram (NAs removed)

```{r}

total.steps.cleaned <- aggregate(as.numeric(cleaned.data$step) ~ as.numeric(cleaned.data$date), data=cleaned.data, sum, na.rm = FALSE)

hist(total.steps.cleaned[,2], xlab="Number of Total Steps", ylab="Frequency (Number of Days)", main="Histogram of Total Observed Steps by Date (NA Values Removed)")


```


### Mean (NAs removed)

```{r}
mean.steps.per.day.cleaned <- mean(total.steps.cleaned[,2])
mean.steps.per.day.cleaned
```


### Median (NAs removed)

```{r}
median.steps.per.day <- median(total.steps[,2])
median.steps.per.day
```


## Are there differences in activity patterns between weekdays and weekends?

```{r}

cleaned.data$date <- as.Date(cleaned.data$date)

day.of.week.test <- weekdays(cleaned.data$date)

days <- c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday")
d.labels <- c("weekday","weekday","weekday","weekday","weekday","weekend","weekend")
cleaned.data$daytype <- factor(day.of.week.test,levels = days, labels = d.labels)


cleaned.averages.by.interval <- aggregate(as.numeric(cleaned.data$step) ~ as.numeric(cleaned.data$interval) + cleaned.data$daytype , data=cleaned.data, mean, na.rm = FALSE)

cleaned.data.split <- split(cleaned.data,cleaned.data$daytype)

weekday.data <- cleaned.data.split$weekday
weekend.data <- cleaned.data.split$weekend

weekday.average.steps.by.interval <- aggregate(as.numeric(weekday.data$step) ~ as.numeric(weekday.data$interval), data=weekday.data, mean, na.rm = FALSE)

weekend.average.steps.by.interval <- aggregate(as.numeric(weekend.data$step) ~ as.numeric(weekend.data$interval), data=weekend.data, mean, na.rm = FALSE)

names(weekday.average.steps.by.interval) <- c("interval","steps")
names(weekend.average.steps.by.interval) <- c("interval","steps")


weekday.average.steps.by.interval$day <- "weekday"
weekend.average.steps.by.interval$day <- "weekend"

combine.back <- data.frame
combine.back <- rbind(weekday.average.steps.by.interval,weekend.average.steps.by.interval)

xyplot(steps  ~ interval | day, data=combine.back, type="l", horizontal = TRUE, layout=c(1,2), ylab="Number of steps", xlab="Interval")


```

